### 2021-05-11 notes/notes.tex
Models back to normality. Before context-weighted adaptation

\begin{tabularx}{\textwidth}{
  l|
  >{\hsize=0.4\hsize\linewidth=\hsize}X|
  >{\hsize=0.2\hsize\linewidth=\hsize}X|
  >{\hsize=0.2\hsize\linewidth=\hsize}X}
End Date & Result & A priori comments & A posteriori comments \\ \hline \hline
30.04.2021 & Lit. research on savings and deadaptation & Are there many with contextual cues? & \\ \hline
07.05.2021 & Finding model pars to explain observations & This is mostly already done &  \\\hline
14.05.2021 & Looking into error-clamp behavior & Maybe there's something to it. If there is, an extra week should be allocated to simulations. &  \\\hline
--- & Start writing the paper. & & \\ \hline

@@

\cite{Huang_Persistence_2009} showed that if participants are placed in an
adaptation environment where the force gradually ramps up (across many trials),
participants show slower de-adaptation during an error-clamp phase, compared to
participants in which the force was suddenly increased during adaptation. They
also showed that participants alter their learning rates to better fit the
volatility of the environment, regardless of the direction of adaptation.
\end{chapter}

@@

\section{The model}
The model is an implementation of \ref{subsec:proposal}. The generalized state
is given by:
\begin{equation}
z_t = \{x, \theta_t, \gamma_t, \omega_t\}
\end{equation}
where $x$ is the current position of the hand in Cartesian coordinates,
where the origin is the starting point. We assume that motor commands are
issued every $\Delta t$ for simplicity. At the beginning of each time interval,
the context is inferred combining the cue and the prediction error of the
outcome of the previous motor command. We assume further than the cue and the
force field are uncorrelated, which yields:
\begin{align}
  q(C_t) &= q(C_t | \omega_t)q(C_t | s_t, s_{t-1}, a_{t-1}) \\ \label{eqn:estimated-context}
  q(C_t | \omega_t) &\propto p(\omega_t | C_t)p(C_t) \\
  q(C_t | x_t, x_{t-1}, a_{t-1}) &\propto p(x_t | C_t, x_{t-1}, a_{t-1})p(x_t)p(x_{t-1})p(C_t)
\end{align}
The terms $p(C_t)$ refer to the prior probability of the context at
the beginning of the trial: at the first trial, this refers to prior beliefs
over which contexts are more common/likely. At each subsequent trial, it
incorporates the belief that has so far been accumulated, given previous
trials.

@@

@@

@@

\subsection{Parameter values to look into}
By visual inspection of the results of simulations, these are the parameter values that seem promising:

\begin{tabularx}{\textwidth}{
  >{\hsize=.2\hsize\linewidth=\hsize}X|
  >{\hsize=.1\hsize\linewidth=\hsize}X|
  >{\hsize=.1\hsize\linewidth=\hsize}X|
  l}
% {l|{\hsize=.4\hsize}X|{\hsize=0.3\hsize}X|{\hsize=0.3\hsize}X}
Agent & Obs noise & Cue noise & Notes \\ \hline \hline
LR    & 1.2       & 0.04      & Slow fake-context inference. Maybe the priors coming from the previous stage are messing things up. \\ \hline
1 & 1 & 1 & 1
\end{tabularx}
>>>>>>> adc91ff996e45d6eaf8974cd33ea69b5d380ed07

\subsection{Adaptation}
Adaptation happens mostly very quickly for all agents. For the current $\Delta t = \dt$, adaptation happens within 20 to 30 time steps, although many times (depending on agent and parametrization) it happens much more quickly. For example, for the LR agent with prediction noise of 0.15, cue noise and context noise of 0, and assumed force variability of 0.25, within 15-20 time steps the adaptation is complete.
\section{Conclusions}

>>>>

\begin{tabularx}{\textwidth}{
  l|
  >{\hsize=0.4\hsize\linewidth=\hsize}X|
  >{\hsize=0.2\hsize\linewidth=\hsize}X|
  >{\hsize=0.2\hsize\linewidth=\hsize}X}
End Date & Result & A priori comments & A posteriori comments \\ \hline \hline
30.04.2021 & Lit. research on savings and deadaptation & Are there many with contextual cues? & I found a lot of evidence. Moving on.\\ \hline
07.05.2021 & Finding model pars to explain observations & This is mostly already done &  \\\hline
14.05.2021 & Looking into error-clamp behavior & Maybe there's something to it. If there is, an extra week should be allocated to simulations. &  \\\hline
--- & Start writing the paper. & & \\ \hline

@@

\cite{Huang_Persistence_2009} showed that if participants are placed in an
adaptation environment where the force gradually ramps up (across many trials),
participants show slower de-adaptation during an error-clamp phase, compared to
participants in which the force was suddenly increased during adaptation. They
also showed that participants alter their learning rates to better fit the
volatility of the environment, regardless of the direction of adaptation.

\section{Notes on savings, quick deadaptation and contextual cues}
In \citep{Kojima_Memory_2004}, Figure 1a-b, we can see that de-adaptation was a
bit faster than adaptation for the gain-increase experiment, and much faster
for the gain-decrease experiment. It seems like there is no contextual cue
available to the monkey, as all trials (over 1k) were presented one after the
other with no breaks. Poor monkey.

\citep{Kojima_Memory_2004} additionally show savings in the form of
re-adaptation that, for a few hundred trials, is much faster than the original
adaptation (control adaptation). This can be better seen in Figure3a-b. They
discuss that there is even a point of inflection in test-adaptation that
clearly marks the end of savings and resuming of adaptation. The duration of
the savings phase changed from experiment to experiment, from monkey to
monkey. The authors do not check whether the past-facilitation rate of learning
is similar to that of the end of the control adaptation phase, which would
indicate that the monkey quickly went back to where it had left off.

Finally, they show that savings seem to disappear after a period of
error-clamp-like trials, in which the ISS is set to zero, thus allowing the
monkey hundreds of trials in which his adapted saccades (gain != 1) yielded
correct results. In contrast, savings do not disappear after the monkey head
was covered with a black cloth, placing him in total darkness (no feedback for
his saccades).

Given the lack of contextual cues, context identification could be very slow in
monkeys, which would explain the facilitation period. During normometric
saccades after adaptation (error-clamp-like trials), context inference could
have been cimented to the point that it became difficult for the monkey to
shift away from it once test adaptation began, thus removing savings from the
equation. During darkness trials, in contrast, there is no contextual evidence
in any direction, and thus the precision over context does not change. In this
case, savings are back on the table.

\cite{Ethier_Spontaneous_2008} present a simple experiments with saccade
adaptation in humans. In their paradigm, humans go through a O-A-B-E paradigm,
in which the adaptation is forced by making the target move closer or further
away than it was before, making participants believe that they had
over/under-shot the target. While this study does not quite talk about savings,
there is evidence of context-based savings in their paradigm, as adaptation
trials were done in blocks of 60 trials, with 30s pauses between blocks, in
which participants closed their eyes. In their figure 2, they show that each
pause caused a dip in the observed adaptation. While this pause does not
provide any clues as to the identity of the following block, it does indicate
that a new block will start, which could make participants revert to using
their priors over contexts to make the first decision in the following block
and consequently using the feedback to perform context inference, explaining
the quick dip after each pause.

Their study does show quick de-adaptation during a second block of trials in
which the desired gain is opposite of the adaptation trials. However, it is
difficult to interpret these savings in terms of the context-based learner, as
learning in the original adaptation is also quite fast, and the B context had
not been introduced in advance. However, the fact that group 2 (which has
longer pauses) shows faster de-adaptation supports the idea of context-based
learning, as the longer pause before de-adaptation causes context inference to
rely more on priors and thus make the jump from model A to model O almost
immediately, and then summoning model B and learning at the same speed as model
A was learned.

Importantly, the effect of these pauses speak for continuity-based priors,
which are based on previous trials and not on prior beliefs regarding which
contexts would, in general, be encountered more often in the world.

\cite{Huang_Persistence_2009} presents interesting data, but I can't interpret
it for now.

\cite{Lee_Dual_2009} present a study that is very useful to us. In it,
participants have to use a joystick to move a cursor from its initial position
to a target, under different visuomotor rotations (tasks A, B, C and D). Of
special significance is the fact that participants were provided with very
clear contextual cues, as the target area was in a position that uniquely
identified the rotation. The effects of these cues can be seen in their figure
6, in which they show how participants very quickly learn the adaptations
during an A-B-A phase, with very quick jump from A to B and from B to A. Then,
participants enter a phase in which tasks C and D are randomly chosen every
trial. Because of the unambiguity of the contextual cues, even during learning
participants display very clear switching of internal models. This is so far
the most clear example of contextual cues causing immediate ``deadaptation''
(which we know is really switching, but don't tell anybody).

\cite{Forano_Timescales_2020} present a study with humans and a mechanical arm,
in which participants have to perform forward reaching movements under
different forces. During their experimental sessions, their complications for
which are endless and unnecessary, there is a baseline phase, in which
participants perform no-force-field movements in two different workspaces
(strong contextual cues). Then an adaptation phase starts, in which
participants do eight trials in context A, eight in B (with random Es thrown in
for good measure). Awful things happen later on, of which we do not
speak. During de-adaptation, they follow the same 8-8 structure, but the cues
are flipped; this phase ends when the adaptation is taken back to zero. Finally
the error-clamp trials start, with a contextual cue that is the same as the
first one they got during the adaptation phase. They do not speak of savings,
and savings cannot be seen in their plot, but quick de-adaptation is clearly
seen in their figures 6 and 7. However, it is not as quick as it could be, and
I believe this to be due to the strength of the cues, which cause a lot of
conflict in the participant as the force-feedback and contextual cues are so at
odds with each other. I would argue that the slow decay (still fast compared to
adaptation) is evidence of yet another phenomenon: the devaluation (or
re-evaluation) of contextual cues, in which the precision with which they are
treated decays over time as a consequence of the apparently false information
they provide. I expect this to interact with the following error-clamp trials,
in which \cite{Forano_Timescales_2020} show small-but-significant spontaneous
recovery. I believe that if those de-adaptation trials were removed, the
increased strength of the contextual cues would make the spontaneous recovery
larger.

\cite{Davidson_Scaling_2004} tested quick de-adaptation more thoroughly than is
usual, with the explicit goal of checking whether there was something special
about A-O de-adaptation that made it quick, or if A-B de-adaptation would also
be quick. Additionally, they explored the question of which one is faster. They
used the typical curling-force mechanical arm for their experiment with human
participants. Their results can be groupped into two: (1) De-adaptation from A
to O is slower than de-adaptation from A to 1/3A. (2) De-adaptation from A to
3A is perhaps a bit faster than from A to -A, but, more importantly,
re-adaptation to A is faster coming from 3A than from -A. The first result can
be explained simply: de-adapting from A to A/3 implies less ``travel''
(i.e. adaptation) than from A to 0, and thus motor error takes longer to go
down when de-adapting to O. This is precisely why \cite{Davidson_Scaling_2004}
performed their second experiment from which (2) was obtained. Their second
result can be explained by switching internal models, if action selection is
done via a weighted sum of the actions generated by each possible model (where
the weights are the posterior probabilities of the models). In effect, the
assymetry is caused by the existance of O. In this account, going from 3A to A
more quickly reduces motor error because the contributions of 3A and O to motor
output ``cancel out'', as they point in opposite directions relative to A,
leaving only the contributions of A. When coming from -A, the contributions of
-A and O point in the same (incorrect) direction (relative to the real
generating process A), thus making motor error be bigger.

Re-adaptation in \cite{Davidson_Scaling_2004} differs greatly between -A and
2A. While the explanation above could be enough, I believe that an extra factor
would be the fact that while adapting to forces in different directions,
different muscles must be used. The transfer from using a model that uses a set
of muscles to another that uses another set of muscles might come with its own
perils. More research into this could be useful.

Interestingly, \cite{Davidson_Scaling_2004} calculated that re-adaptation from
3A to A was not significantly different in experiments 1 and 2, even though
participants in experiment 2 had already learned the A environment, while those
in experiment 1 had not. This can perhaps be explained as a fluke in the
switching mechanism, where the weight of the A model in experiment 2 made its
contribution to action lie square in the middle of those from 3A and O. This
might need further consideration.

\cite{Zarahn_Explaining_2008} performed a typical experiment with visuomotor
modifications (screen, hand position, axis rotation) to show that in their
O-A-W-A trial, where W are washout trials, savings still existed, even after
the fast-slow learner predicted that savings should have disappeared. They
showed the same with O-A-(-A)-A and who knows what else. Their figure 3 is a
nice poster child. Note that while they did not have any contextual cues, their
transformations are HUGE. In their O-A-W-A experiment, A = 45°. A = 30° in their
other experiment. This in stark contrast with many other studies that use
10°-20° ranges.

\cite{Kim_Neural_2015} present savings with very informative and clear
cues. Their savings are beautifully portraited in their figure
2A. Additionally, they present imaging results based on regressors obtained
from the fast-slow linear models.

\section{Notes on the effects of context inference on learning}

\cite{Herzfeld_memory_2014} present what is possibly (so far) the most
compelling evidence for the effects of contex-inference on learning itself,
uncoupled from motor responses. The authors make the point (with a model and
experiment) that the brain adjusts its sensitivity to errors based on the
volatility of the environment. Moreover, this sensitivity depends on the size
of the error, such that a participants' learning might be faster when facing
small errors than with large ones. In their first experiment, the authors
divided participants into three groups that differed only on the volatility of
the environment, group 1 facing the most stable environment, where the
probability that the perturbation changed from 13 to -13 (N s/m) was 0.1; group
2 had a probability of 0.5 and group3 of 0.9. \cite{Herzfeld_memory_2014}
labeled these groups as 0.9, 0.5 and 0.1, respectively. Here I keep with these
labels.

\cite{Herzfeld_memory_2014} showed in their figure 1D that the stable
environment of group 0.9 made participants learn more quickly from errors, with
the effects being lower for groups 0.5 and 0.1, in that order. In my model, the
interpretation of these results is fairly straightforward: more volatile
environments lead to less precise (and more often incorrect) context inference,
which would slow down learning. In particular, in the mathematical
implementations of the model so far (see \ref{sec:model-instance}), the
precision with which observations are taken in depends on the precision of the
context inference, which should create behavior like that observed by
\cite{Herzfeld_memory_2014}. Note that the authors' results in experiment 2 are
explained by my model in very much the same vein.

One puzzling result from \citep{Herzfeld_memory_2014} is that sensitivity to
error depends on the size of the error. With some effort (invested in
understanding their paper...), this can be very nicely explained by the
context-inference model. First, to put some of their numbers in perspective,
their second experiment implies errors roughly in the range of -2 to 2 (which
matches the x-axis in Figure 2). An error the size of -2 or 2 very clearly
signals the change from one context to the other (0.9 gain to 1.1, or vice
versa; it is unclear in the paper). Because the direction of the change is
clear evidence from the previous and current contexts, context inference right
after such a switch becomes very clear (the maximum of the posterior $>>$ the
minimum of the posterior), which makes learning somewhat fast and does not
depend on the volatility of the environment, at least in the first trial after
a switch; during the next trials, the uncertainty increases again due to
volatility. The smaller the error is (in absolute value), the less clearly it
can be attributed to a switch in the environment, to the extreme where changes
are minuscule and thus must be attributed to the motor adaptation's error; it
is because of this that small errors have a higher effect on learning. Now, for
smaller errors, there is a clear effect of volatility on learning. I believe
this can be explained by the context-learner: Because of the overall slower
learning in volatile environments, the internal models are, for any given
trial, not as well learned for the high-volatility groups, which leads to many
errors being misinterpreted as context switches (for which no learning might
occur), making the average look worse than it is (except, as explained above,
when the error is close to 2 or -2, which makes switches very evident).

In fact, this introduces important confounds: errors of exactly 2 and -2 can
only ocurr when the participant has learned both internal models perfectly and
simply predicted the wrong one, thus adapting in the incorrect direction. In
this case, learning should be exactly 0. However, errors very close to, but not
exactly, 2 and -2 can be attributed to bad context inference plus the need to
adapt the ``other'' model, i.e. the model that was not predicted. In this case,
depending on how smart the participant is, the other model might be updated
with an error sensitivity as high as that of the near zero error. On the other
hand, participants observing an error close to zero might interpret this as a
signal to update an internal model (in which case the sensitivity should be
high), but they might also falsely (or even correctly) accuse the context of
having changed, but with low precision, making learning of the ``other'' model
very low (due to low precision on context), or even zero, if the participant is
so unsure that she does not want to learn from this. Because of this, a more
correct version of Figure 2E would take into account how well the internal
models have been learned at each trial, how many trials were spent in
perfect-model land, how many times the participant took a near-2 error as a
switch (correctly and incorrectly), how many times near-2 errors occurred when
the models been already fully learned, etc.

The third experiment by \cite{Herzfeld_memory_2014} has too many
confounds. They conclude that they can independently manipulate sensitivity to
big and small errors, but I believe their experimental design has unwanted
consequences because 8N and -8N were shown a lot during the fast-switch part,
but 4N and -4N were not. It's such a mess that I don't want to touch it.

\section{Notes on error-clamp weirdness}
\cite{Kojima_Memory_2004} have error-clamp-like trials, but not at the end. No
weirdness here. Also no cues.

\cite{Smith_Interacting_2006} speak of error-clamp weirdness and presents a
model to explain it. Their data is limited, however, and might not be specially
useful for me.

\cite{Ethier_Spontaneous_2008} have error-clamp weirdness and their
within-block pauses might be useful for my interpretations.
\end{chapter}

\cite{Forano_Timescales_2020} has spontaneous recovery and slow tapering off
during error-clamp trials, but it's quite small. Maybe because of the large
de-adaptation that devalues contextual cues?

\cite{Vaswani_Decay_2013} explores in detail the motor decay observed during EC
trials after adaptation trials. They conclude, among other things, that the
motor adaptation decays across as many as hundreds of trials, but never goes
back to zero. Additionally, there is a lag before decay starts (all their
experiments end in A-E, so no baseline before), and it depends on the
participant, as well as how noticeable the transition to error-clamp trials
is. They have some clever experiments.

\section{Notes on implicit vs explicit, and strategies.}
The idea behind strategies in motor adaptation is simple: participants can be
told how to perform in a novel environment. Then, when the novel environment
kicks in, there is no period of adaptation before performance is near-perfect.

For example, \cite{Taylor_Flexible_2011,Mazzoni_Implicit_2006} showed that in a
visuomotor rotation task, participants can be given the hint to ``do as if you
were aiming for this other target'', which allows them to complensate almost
perfectly from trial 1. However, the authors also found interesting effects:
while at first participants have zero error, they slowly drift away from this,
overcompensating for the visuomotor rotation. The authors showed that this
overcompensation goes away, given sufficient trials.

\cite{Taylor_Flexible_2011} present a computational model based on the linear,
prediction-error based models \citep[e.g.][]{Smith_Interacting_2006} by making
prediction error incorporate some element of the strategy. This model is fine
and explains the observed phenomena, but I believe that they overlooked the
mismatch between propriosceptive and visual feedback. This mismatch, which is
independent of reward, would explain their results (at least qualitatively). I
would speculate that the effect of the strategy given to the participants by
\cite{Taylor_Flexible_2011} makes them not adapt their motor rotations, but
rather shift the goal of the trial. Then, as trials go by, the propiosceptive
feedback does not match what is observed on the screen, and that causes motor
adaptation. Perhaps more recent studies have tested this.

@@

\section{The model}
\label{sec:model-instance}
The model is an implementation of \ref{subsec:proposal}. The generalized state
is given by:
\begin{equation}
z_t = \{x, \theta_t, \gamma_t, \omega_t\}
\end{equation}
where $x$ is the current position of the hand in Cartesian coordinates,
where the origin is the starting point. We assume that motor commands are
issued every $\Delta t$ for simplicity. At the beginning of each time interval,
the context is inferred combining the cue and the prediction error of the
outcome of the previous motor command. We assume further than the cue and the
force field are uncorrelated, which yields:
\begin{align}
  q(C_t) &= q(C_t | \omega_t)q(C_t | s_t, s_{t-1}, a_{t-1}) \\ \label{eqn:estimated-context}
  q(C_t | \omega_t) &\propto p(\omega_t | C_t)p(C_t) \\
  q(C_t | x_t, x_{t-1}, a_{t-1}) &\propto p(x_t | C_t, x_{t-1}, a_{t-1})p(x_t)p(x_{t-1})p(C_t)
\end{align}
The terms $p(C_t)$ refer to the prior probability of the context at
the beginning of the trial: at the first trial, this refers to prior beliefs
over which contexts are more common/likely. At each subsequent trial, it
incorporates the belief that has so far been accumulated, given previous
trials.

@@

The effect of the hyperparameters of the priors are worth a note or two, as they are complex.

Naturally, $\mu$ affects the initial estimate of the adaptation, in the same
units as the necessary adaptation. $\nu$ encodes how stable this hyperprior is:
higher values (e.g. 10000) all but guarantee that the hypermean will not move
in the face of evidence; In principle, enough evidence should still move it,
but that won't happen during an experiment. Smaller values (e.g. 1 / force\_sd,
as is the default on the code) make the hypermean (and thus the mean) follow
evidence more freely. Note that as more observations are accumulated, $\nu$
becames bigger and bigger, solidifying the value of the hypermean.

The hyperparameters $\alpha$ and $\beta$ are a bit more complex. Note that the
mean of a Gamma distribution is $\beta / (\alpha \nu)$; this mean is being used
as the standard deviation of a Gaussian by the rest of the agent (cheating),
which makes it an important measure of uncertainty. While setting the default
hyperparameters, the values used are $\alpha = 0.5 / \sigma_0$ and
$\beta = 0.5$, where $\sigma_0$ is the \textit{a priori} estimate of the
standard deviation of the force exerted by the environment (force\_sds in the
code), which controls the initial learning rate. This makes the initial
standard deviation equal $\sigma_0$, which makes it consistent with the
fixed-force model in its interface. The 0.5 values ensure that uncertainty is
large at the beginning and during the experiment is greatly reduced, but never
to a point where it's ``visually'' too small. Changing this 0.5 would make the
standard deviation change more quickly, making the model more or less precise
in its predictions, independently of the volatility of the mean of the
adaptation (via the hypermean).

The baseline model defaults to different values that make it a lot more
stable. The hyperstd of the mean is set to 10,000, which makes the mean
entirely stable during the duration of the experiment. The values of $\alpha$
and $\beta$ are fixed regardless of $\sigma_0$ such that the standard deviation
is 0.001 (compare that to the size of the adaptations, around 0.0125), and the
hyperparameters of the standar deviation are stable during the experiment.

@@

\subsubsection{Example: LR model with bad cues}
Using the LR models with only two contexts (baseline, red) creates slow
adaptation (across 400 trials or so), with fast-but-not-immediate switch back
to baseline when provided with false cues: after the adaptation block ends, a
baseline block ends but the cue still points to the adaptation block.

Simulations from this can be seen in the function
slow\_context\_inference\_badcues of /adaptation/code/simulations.py

The agent parameters are as follows:

\begin{center}
\begin{tabularx}{0.8\textwidth}{ c | c | X}
  Variable & Value & Description \\ \hline
  prediction\_noise & 0 & Setting this to a nonzero value screws everything up. May be a bug. \\  
  cue\_noise & 0.015 & Changing this by 0.001 creates slower or faster context inference \\
  force\_sds & 0.0002 & Changing this makes the model learn more quickly or slowly. With this value, the agent barely finishes adaptation by the time the adaptation block is over. \\
  context\_noise & 0 & Higher numbers make context inference more uncertain after evidence accumulation.
\end{tabularx}
\end{center}

\subsubsection{Example: LR model with no cues}
Using the LR models with only two contexts (baseline, red) creates slow
adaptation (across 400 trials or so), with fast-but-not-immediate switch back
to baseline when no cues are provided. After each block (A or O), a slow switch
can be observed, which is based entirely on the internal models producing noisy
predictions.

Simulations from this can be seen in the function
slow\_context\_inference\_badcues of /adaptation/code/simulations.py

\begin{center}
\begin{tabularx}{0.8\textwidth}{ c | c | X}
  Variable & Value & Description \\ \hline
  prediction\_noise & 3 & The bigger this is, the slower correct inference is. Only with no cues, though. \\  
  cue\_noise & 0.5 & Creates uninformative cues.
  force\_sds & 0.0002 & Changing this makes the model learn more quickly or slowly. With this value, the agent barely finishes adaptation by the time the adaptation block is over.
  context\_noise & 0 & Higher numbers make context inference more uncertain after evidence accumulation.  
\end{tabularx}
\end{center}

@@

\section{Conclusions}
